Stat 540 Seminar 3
========================================================

The purpose of this document is to explore and learn ggplot2 plotting functions.

load libraries subsample data:
```{r}
library(ggplot2)

prDat <- read.table("../GSE4051_data.tsv")
prDes <- read.table("../GSE4051_design.tsv", header = TRUE)

set.seed(540) #for consisteny

#sample random genes and reshape
prDat2 <- t(prDat[sample(1:nrow(prDat), size = 2),])
genes <- colnames(prDat2)
prDat2 <- data.frame(sidChar = rownames(prDat2), prDat2, check.names=FALSE) #names for merging
prDat2 <- merge(prDes,prDat2)
oDat2 <- with(prDat2,data.frame(sidChar, sidNum, devStage, gType, probeset = factor(rep(genes, each = nrow(prDat2))), geneExp = unlist(prDat2[,genes])))
```


### Stripplots

Then let's plot the expression level of each gene.
```{r}
p <- ggplot(oDat2, aes(geneExp, probeset)) + geom_point(position = position_jitter(height = 0.1))
p
```

Now let's explore gene expression changes over the course of development.
```{r}
p <- p <- ggplot(oDat2, aes(devStage, geneExp)) + geom_point()
p <- p + facet_wrap(~ probeset)     
p <- p + aes(color = gType)
p <- p + stat_summary(fun.y = mean, geom = "point", shape = 4, size = 4)
p
```

### Density plots

Lets use density plots to investigate the distribution of gene expression in these 2 genes.

```{r}
p <- ggplot(oDat2, aes(geneExp, color = gType)) + stat_density(geom = "line", position = "identity") + geom_point(aes(y = 0.05), position = position_jitter(height = 0.005))
p
```

### Box plots and violin plots

Lets investigate the distribution of gene expression across developmental stages.

```{r}
p <- ggplot(oDat2, aes(devStage, geneExp)) + 
   geom_boxplot()
p <- p + facet_wrap(~ gType)
p
```

Now with a violin plot

```{r}
p <- ggplot(oDat2, aes(devStage, geneExp)) + 
   geom_violin()
p <- p + facet_wrap(~ gType)
p
```

### Plot matrix

Sample some data to investigate how to deal with over plotting

```{r}
#pairDat <- subset(prDat, select = sample(1:ncol(prDat), size = 4))
```

Plot Data using binhex to show density of the points.

```{r}
#p <- p <- plotmatrix(pairDat) + stat_binhex()
#p
```

### HeatMap in ggplot2

Now lets use ggplot to plot a heatmat to deal with a larger dataset.

Start by sampling 20 genes and shape data for clustering:
```{r}
set.seed(540) #for consisteny
prDat20 <- t(prDat[sample(1:nrow(prDat), size = 20),])

hDat <- as.matrix(prDat20)
rownames(hDat) <- with(prDes, paste(devStage, gType, sidChar, sep="_"))

# genes <- colnames(prDat20)
# prDat20 <- data.frame(sidChar = rownames(prDat20), prDat20, check.names=FALSE) #names for merging
# prDat20 <- merge(prDes,prDat20)
# hDat20 <- with(prDat20,data.frame(y = paste(devStage, gType, sidChar, sep="_"), x = factor(rep(genes, each = nrow(prDat20))), geneExp = unlist(prDat20[,genes])))

#for dedrogram
library(ggdendro)

library(reshape2)

#prepare dedrogram clustering
dd.col <- as.dendrogram(hclust(dist(hDat)))
col.ord <- order.dendrogram(dd.col)

dd.row <- as.dendrogram(hclust(dist(t(hDat))))
row.ord <- order.dendrogram(dd.row)

ddata_x <- dendro_data(dd.row)
ddata_y <- dendro_data(dd.col)

xx <- prDat20[col.ord, row.ord]
xx_names <- attr(xx, "dimnames")
df <- as.data.frame(xx)
colnames(df) <- xx_names[[2]]
df$gene <- xx_names[[1]]
df$gene <- with(df, factor(gene, levels=gene, ordered=TRUE))

hDat_gene <- melt(df, id.vars="gene")

```

Plot:
```{r }
#for custom pallette
library(RColorBrewer)
jBuPuFun <- colorRampPalette(brewer.pal(n = 9, "BuPu"))

#for proper formatting
library(grid)

p1 <- ggplot(hDat_gene, aes(x=gene, y=variable)) + geom_tile(aes(fill=value))
p1 <- p1 + theme(axis.text.x=element_text(angle=90, hjust=1))
p1 <- p1 + scale_fill_gradientn(colours = jBuPuFun(100))

theme_none <- theme(
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  axis.title.x = element_text(colour=NA),
  axis.title.y = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.line = element_blank()
)

# Dendrogram 1
p2 <- ggplot(segment(ddata_x)) + geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) + 
  theme_none + theme(axis.title.x=element_blank()) + coord_flip()

# Dendrogram 2
p3 <- ggplot(segment(ddata_y)) + geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) + theme_none

grid.newpage()
print(p1, vp=viewport(0.9, 0.9, x=0.4, y=0.4))
print(p2, vp=viewport(0.2, 0.81, x=0.9, y=0.46))
print(p3, vp=viewport(0.665, 0.2, x=0.415, y=0.92))
```

I'll say that although it is possible to do in ggplot, the combined dendrogram plot should be done in lattice. Heatmaps by themselves can be easily done in ggplot however.
Method was adapted from: http://stackoverflow.com/questions/6673162/reproducing-lattice-dendrogram-graph-with-ggplot2/6675983#6675983



