Stat 540 Seminar 2
========================================================

The purpose of this document is gain an understanding of probablility distributions. This document is based on Jenny's version of the seminar [tutorial](http://www.ugrad.stat.ubc.ca/~stat540/seminars/seminar02_playing-with-probability.html).

Start by generating data:
```{r}
n <- 10
B <- 4

set.seed(540) # to obtain deterministic results
x <- matrix(rnorm(n*B), nrow = n)

#check the data
str(x)
head(x)
```

Jenny's tutorial goes through other methods for generating data but this the nicest R-like way of doing it.

Give that data row and column names
```{r}
rownames(x) <- sprintf("obs%02d", 1:n)
colnames(x) <- sprintf("samp%02d", 1:B)
x #check what this looks like
colMeans(x) #fast way compute the means for all the samples
```

### Exercise: Sample mean vs true mean
```{r}
mean(x) #true mean
mean(colMeans(x)) #sample mean
```

They seem to be identical. I will try with another seed and n, b value:
```{r}
n <- 10
B <- 4
set.seed(77) # to obtain deterministic results
x <- matrix(rnorm(n*B), nrow = n)
colMeans(x)
mean(x) #true mean
mean(colMeans(x)) #sample mean
```

Same results again. I expected them to be so, given the equal number of elements used to compute each sample mean.

### Exercise: Explore the Weak Law of Large Numbers

Note SEM mean standard error of the mean

```{r}
B = 1000
#generate data
n10 <- colMeans(matrix(rnorm(10*B), nrow = 10))
n100 <- colMeans(matrix(rnorm(100*B), nrow = 100))
n1000 <- colMeans(matrix(rnorm(1000*B), nrow = 1000))
n10000 <- colMeans(matrix(rnorm(10000*B), nrow = 10000))

lawDat <- data.frame( sampSize = c(10,100,1000,10000), trueSEM = 1/sqrt(c(10,100,1000,10000)), obsSEM = c( sd(n10), sd(n100), sd(n1000), sd(n10000)), sampMeanIQR = c(IQR(n10), IQR(n100),IQR(n1000),IQR(n10000)), sampleMeanMad = c(mad(n10),mad(n100),mad(n1000),mad(n10000)))

lawDat
```

### Exercize: Repeat with alternate distribution

```{r}
#generate data
n10 <- colMeans(matrix(rnorm(10*B, sd = 2), nrow = 10))
n100 <- colMeans(matrix(rnorm(100*B, sd = 2), nrow = 100))
n1000 <- colMeans(matrix(rnorm(1000*B, sd = 2), nrow = 1000))
n10000 <- colMeans(matrix(rnorm(10000*B, sd = 2), nrow = 10000))

lawDat <- data.frame( sampSize = c(10,100,1000,10000), trueSEM = 1/sqrt(c(10,100,1000,10000)) * 2, obsSEM = c( sd(n10), sd(n100), sd(n1000), sd(n10000)), sampMeanIQR = c(IQR(n10), IQR(n100),IQR(n1000),IQR(n10000)), sampleMeanMad = c(mad(n10),mad(n100),mad(n1000),mad(n10000)))

lawDat
```

### Generate a large sample from some normal distribution and explore probabilities with respect to relative frequencies

```{r}
normSamp <- rnorm(10000)
#pick a threshold lets say 0.5 sd from the mean
pnorm(0.5)
length(normSamp[normSamp <= 0.5])/10000
```

Pretty close. Lets try another distribution

```{r}
samp <- rnorm(10000, sd = 2)
#pick a threshold lets say 0.5 sd from the mean
pnorm(0.5, sd = 2)
length(samp[samp <= 0.5])/10000
```

Lets try a smaller sample size

```{r}
samp <- rnorm(100, sd = 2)
#pick a threshold lets say 0.5 sd from the mean
pnorm(0.5, sd = 2)
length(samp[samp <= 0.5])/100
```

It seems to drift away from the true results. Lets try probabilites greater than a threshold.

```{r}
samp <- rnorm(10000, sd = 2)
#pick a threshold lets say 1.1 sd from the mean
pnorm(1.1, sd = 2, lower.tail = FALSE)
length(samp[samp > 1.1])/10000
```

Now lets try a interval between 2 thresholds.

```{r}
samp <- rnorm(10000, sd = 2)
#pick a threshold lets say -0.5 and 1.1 sd from the mean
pnorm(1.1, sd = 2) - pnorm(-0.5, sd = 2)
length(samp[samp <= 1.1 & samp >= -0.5])/10000
```

Now lets try a interval between 2 thresholds.


```{r}
samp <- rnorm(10000, sd = 2)
#pick a threshold lets say -0.5 and 1.1 sd from the mean
pnorm(1.1, sd = 2) - pnorm(-0.5, sd = 2)
length(samp[samp <= 1.1 & samp >= -0.5])/10000
```

### Recreation of Jenny's lattice plot with ggplot

```{r}
library(ggplot2)
library(lattice)

B <- 1000
n <- round(10^(seq(from = 1, to = 2.5, length = 4)), 0)
names(n) <- paste0("n", n)
getSampleMeans <- function(n, B) colMeans(matrix(rnorm(n * B), nrow = n))
x <- data.frame(sapply(n, getSampleMeans, B))
jFormula <- as.formula(paste("~", paste(names(n), sep = "", collapse = " + ")))
xTallSkinny <- stack(x)
names(xTallSkinny) <- c("x","n")
xTallSkinny$n <- factor(xTallSkinny$n, levels = colnames(x))


plot <- ggplot(xTallSkinny, aes(x, group = n, colour = n)) + geom_density()
plot <- plot + geom_point( aes(x = x, y = 0, group = n, colour = n), position = "jitter", pch = 1)
plot <- plot + xlab("Sample Means")
plot

densityplot(~ x, xTallSkinny, xlab = "sample means", groups = n,
            auto.key = list(x = 0.9, y = 0.9, corner = c(1, 1),
                            reverse.rows = TRUE))
```