Stat 540 Seminar 4
========================================================

Purpose: 

Start by loading data/libraries:
```{r}
library(ggPlot2)
library(plyr)
prDat <- read.table("../GSE4051_data.tsv")
prDes <- read.table("../GSE4051_design.tsv", header = TRUE)
```

### Editing Inner Function to use t test, Wicoxon Test and KS test

Let's extract 6 genes at random, rather than hardcoded genes:
```{r}
set.seed(540)
keepGenes <- row.names(prDat)[sample(1:nrow(prDat), 6)]
miniDat <- subset(prDat, rownames(prDat) %in% keepGenes)
miniDat <- data.frame(gExp = as.vector(t(as.matrix(miniDat))), gene = factor(rep(rownames(miniDat), each = ncol(miniDat)), levels = keepGenes))
miniDat <- suppressWarnings(data.frame(prDes, miniDat))
```

Next lets define some functions for running each test
```{r}
#for t test
tTest <- function(z) {
  zz <- t.test(gExp ~ gType, z)
  round(c(tStat = zz$statistic, pVal = zz$p.value), 4)
}

#for Wilcoxon test
wilTest <- function(z) {
  #not sure if suppressing warnings is the way to go but the error messages are annoying
  suppressWarnings(zz <- wilcox.test(gExp ~ gType, z))
  round(c(wilStat = zz$statistic, pVal = zz$p.value), 4)
}

#for KS test
ksTest <- function(z) {
  suppressWarnings(zz <- ks.test(z[z$gType == "wt", "gExp"], z[z$gType == "NrlKO", "gExp"]))
  round(c(ksStat = zz$statistic, pVal = zz$p.value), 4)
}

#for all 3 tests
allTest <- function(z) {
  zzT <- t.test(gExp ~ gType, z)
  suppressWarnings(zzWil <- wilcox.test(gExp ~ gType, z))
  suppressWarnings(zzKS <- ks.test(z[z$gType == "wt", "gExp"], z[z$gType == "NrlKO", "gExp"]))
  round(c(pVal_T = zzT$p.value, pVal_Wil = zzWil$p.value, pVal_KS = zzKS$p.value), 4)
}
```

Test out the functions:
```{r}
ddply(miniDat, ~ gene, tTest)
ddply(miniDat, ~ gene, wilTest)
ddply(miniDat, ~ gene, ksTest)
ddply(miniDat, ~ gene, allTest)
```

### Matrix of p-values and genes for 100 sampled genes

Start by sampling 100 genes:
```{r}
geneSample100 <- row.names(prDat)[sample(1:nrow(prDat), 100)]

sample100Dat <- subset(prDat, rownames(prDat) %in% geneSample100)

sample100Dat <- data.frame(gExp = as.vector(t(as.matrix(sample100Dat))), gene = factor(rep(rownames(sample100Dat), each = ncol(sample100Dat)), levels = geneSample100))

sample100Dat <- suppressWarnings(data.frame(prDes, sample100Dat))

#write new function to format data in long format
#for t test
tTest <- function(z) {
  zz <- t.test(gExp ~ gType, z)
  data.frame(type = "t", pVal = as.numeric(round(zz$p.value,4)))
}

#for Wilcoxon test
wilTest <- function(z) {
  #not sure if suppressing warnings is the way to go but the error messages are annoying
  suppressWarnings(zz <- wilcox.test(gExp ~ gType, z))
  data.frame(type = "wil", pVal = as.numeric(round(zz$p.value,4)))
}

#for KS test
ksTest <- function(z) {
  suppressWarnings(zz <- ks.test(z[z$gType == "wt", "gExp"], z[z$gType == "NrlKO", "gExp"]))
  data.frame(type = "ks", pVal = as.numeric(round(zz$p.value,4)))
}

pvalDat <- rbind(ddply(sample100Dat, ~ gene, tTest), ddply(sample100Dat, ~ gene, wilTest), ddply(sample100Dat, ~ gene, ksTest))


p <- ggplot(pvalDat, aes(gene, pVal, colour = type)) + geom_point()
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1))
p
```

Not really sure what I am supposed to get out of this, seems chaotic. Lets try the log scale like suggested:

```{r}
(p <- p+ scale_y_log10())
```

A little better, I guess I see that low p-values seems to occur together.

Let's try sorting the data by mean pvalue:
```{r}
#for reordering
pvalDatMean <- ddply(pvalDat, ~gene, summarize, mean = mean(pVal))
pvalDatMean <- pvalueDatMean[order(pvalDatMean$mean), ]

p <- ggplot(pvalDat, aes(x = reorder(gene, pVal), y = pVal, group = type, colour = type)) + geom_point()
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1))+ scale_y_log10()
p
```
The p-values mostly follow the same trend (near the high and low p-values) but seem to be dispersed in the inner regions of the plot.

Now lets investigate what happens when you hard thresholding on p = 0.05:
```{r}
threshDat <- ddply(pvalDat, ~ gene + type, summarize, sig = pVal > 0.95)
sigDat <- ddply(threshDat, ~ gene, summarize, count = length(which(sig)))
```

Significant 0 times:
```{r}
length(which(sigDat$count == 0))
```

Significant 1 times:
```{r}
length(which(sigDat$count == 1))
```

Significant 2 times:
```{r}
length(which(sigDat$count == 2))
```

Significant 3 times:
```{r}
length(which(sigDat$count == 3))
```

### Time limiting study

Plotting how long each test takes. Start by generating samples
```{r}
geneSample5 <- row.names(prDat)[sample(1:nrow(prDat), 5)]
sample5Dat <- subset(prDat, rownames(prDat) %in% geneSample5)
sample5Dat <- data.frame(gExp = as.vector(t(as.matrix(sample5Dat))), gene = factor(rep(rownames(sample5Dat), each = ncol(sample5Dat)), levels = geneSample5))
sample5Dat <- suppressWarnings(data.frame(prDes, sample5Dat))

geneSample10 <- row.names(prDat)[sample(1:nrow(prDat), 10)]
sample10Dat <- subset(prDat, rownames(prDat) %in% geneSample10)
sample10Dat <- data.frame(gExp = as.vector(t(as.matrix(sample10Dat))), gene = factor(rep(rownames(sample10Dat), each = ncol(sample10Dat)), levels = geneSample10))
sample10Dat <- suppressWarnings(data.frame(prDes, sample10Dat))

geneSample50 <- row.names(prDat)[sample(1:nrow(prDat), 50)]
sample50Dat <- subset(prDat, rownames(prDat) %in% geneSample50)
sample50Dat <- data.frame(gExp = as.vector(t(as.matrix(sample50Dat))), gene = factor(rep(rownames(sample50Dat), each = ncol(sample50Dat)), levels = geneSample50))
sample50Dat <- suppressWarnings(data.frame(prDes, sample50Dat))

geneSample500 <- row.names(prDat)[sample(1:nrow(prDat), 500)]
sample500Dat <- subset(prDat, rownames(prDat) %in% geneSample500)
sample500Dat <- data.frame(gExp = as.vector(t(as.matrix(sample500Dat))), gene = factor(rep(rownames(sample500Dat), each = ncol(sample500Dat)), levels = geneSample500))
sample500Dat <- suppressWarnings(data.frame(prDes, sample500Dat))
```

Run Tests:
```{r}

#for ease of returning time
getTime <- function(expr){
  result <- system.time(expr)
  return(result[1])
}

system.time(ddply(sample5Dat, ~ gene, tTest))[1]

timeTTest <- data.frame( time = c(getTime(ddply(sample5Dat, ~ gene, tTest)),
getTime(ddply(sample10Dat, ~ gene, tTest)),
getTime(ddply(sample50Dat, ~ gene, tTest)),
getTime(ddply(sample100Dat, ~ gene, tTest)),
getTime(ddply(sample500Dat, ~ gene, tTest))), testNum = c(5,10,50,100,500), type = "tTest")

timeWilTest <- data.frame( time = c(getTime(ddply(sample5Dat, ~ gene, wilTest)),
getTime(ddply(sample10Dat, ~ gene, wilTest)),
getTime(ddply(sample50Dat, ~ gene, wilTest)),
getTime(ddply(sample100Dat, ~ gene, wilTest)),
getTime(ddply(sample500Dat, ~ gene, wilTest))), testNum = c(5,10,50,100,500), type = "wilTest")

timeKSTest <- data.frame( time = c(getTime(ddply(sample5Dat, ~ gene, ksTest)),
getTime(ddply(sample10Dat, ~ gene, ksTest)),
getTime(ddply(sample50Dat, ~ gene, ksTest)),
getTime(ddply(sample100Dat, ~ gene, ksTest)),
getTime(ddply(sample500Dat, ~ gene, ksTest))), testNum = c(5,10,50,100,500), type = "ksTest")

timeDat <- rbind(timeTTest, timeWilTest, timeKSTest)
```

According to this plot it scales linearly, but it is already reaching >9 seconds for 500 samples for some tests.

Plot the results:
```{r}
p <- ggplot(timeDat, aes(testNum, time, group = type, colour = type)) + geom_line()
p
```