Stat 540 Seminar 6
========================================================

To explore linear models on gene expression data, with limma (higher volume).

Start by loading data/libraries:
```{r}
library(limma)
library(lattice)
library(ggplot2)
prDat <- read.table("../GSE4051_data.tsv")
prDes <- read.table("../GSE4051_design.tsv", header = TRUE)
```

Simulate 100 genes
```{r}
m <- 1000
n <- 3
x <- matrix(rnorm(m * n), nrow = m)
obsVars <- apply(x, 1, var)
summary(obsVars)
mean(obsVars < 1/3)
densityplot(~obsVars, n = 200)
```

## Fit a linear model: explain gene expression in the wild type mice as a function of developmental stage (one-way ANOVA)

```{r}
wtDat <- subset(prDat, select = prDes$gType == "wt")
str(wtDat, max.level = 0)
wtDesMat <- model.matrix(~devStage, wtDes)
str(wtDesMat)
wtFit <- lmFit(wtDat, wtDesMat)
wtEbFit <- eBayes(wtFit)
topTable(wtEbFit)
topTable(wtEbFit, coef = 2:5)
colnames(coef(wtEbFit))
(dsHits <- topTable(wtEbFit, coef = grep("devStage", colnames(coef(wtEbFit)))))
```

## Write a function to prepare a mini-dataset for a small number of genes

```{r}
(luckyGenes <- c("1419655_at","1438815_at"))
jDat <- prepareData(luckyGenes)
str(jDat)
head(jDat)
tail(jDat)
```

## Write a function to stripplot a mini-dataset

```{r}
makeStripplot(jDat)
makeStripplot(jDat, pch = 17, cex = 3)
```

You can use both of your functions together and create a minidatset and plot it all at once:
```{r}
makeStripplot(newDat <- prepareData("1456341_a_at"))
str(newDat)
head(newDat)
```

## Do a two-sample t-test

Here's what I get:
```{r}
mDat <- prepareData("1438786_a_at")
t.test(gExp ~ devStage, mDat, 
       subset = devStage %in% c("P2", "4_weeks"),
       var.equal = TRUE)
```
## Fit a linear model with a categorical covariate

```{r}
mFit <- lm(gExp ~ devStage, mDat, subset = gType == "wt")
summary(mFit)
```

## Perform inference for a contrast

Helpful link for making correct contrast matrix http://www.vsni.co.uk/software/genstat/htmlhelp/anova/Contrasts.htm
```{r}
contMat <- matrix(c(0, 1, 0, -1, 0), nrow = 1)
obsDiff <- contMat %*% coef(mFit)
obsDiff
```

Get p-value for contrast
```{r}
estSe <- contMat %*% vcov(mFit) %*% t(contMat)
testStat <- obsDiff/estSe
2 * pt(abs(testStat), df = df.residual(mFit), lower.tail = FALSE)
```

## Fit a linear model with two categorical covariates

```{r}
makeStripplot(oDat <- prepareData("1448690_at"))
oFitBig <- lm(gExp ~ gType * devStage, oDat)
summary(oFitBig)$coef
oFitSmall <- lm(gExp ~ gType + devStage, oDat)
summary(oFitSmall)$coef
```

Results slighly different due to R picking a differnt intercept to use for some reason. Overall interpretation seems the same.

Peform Anova:
```{r}
anova(oFitSmall, oFitBig)
```

## Apply methods on multiple genes

Sample some data:
```{r}
set.seed(540)
genes12 <- row.names(prDat)[sample(1:nrow(prDat), 12)]
jDat <- prepareData(genes12)
makeStripplot(jDat)
```

Function for performing an anova on categorical covariates devStage and gType:
```{r}
anovaDSGT <- function(x)
  {
  makeStripplot(oDat <- prepareData("1448690_at"))
  oFitBig <- lm(gExp ~ gType * devStage, x)
  oFitSmall <- lm(gExp ~ gType + devStage, x)
  return(as.numeric(anova(oFitSmall, oFitBig)$Pr[2]))
  }

ddply(jDat, ~gene, anovaDSGT)
```

1416357_a_at shows some interaction as well as possibly 1450169_at.

## Use quanitative time variable instead of devStage

```{r}
library(car)

#method with age variable
prepareDataAge <- function(x)
  {
  miniDat <- subset(prDat, rownames(prDat) %in% x)
  miniDat <- data.frame(gExp = as.vector(t(as.matrix(miniDat))), gene = factor(rep(rownames(miniDat), each = ncol(miniDat)), levels = x))
  miniDat <- data.frame(prDes, miniDat)
  miniDat$sidNum <- as.numeric(miniDat$sidNum)
  miniDat$age <- suppressWarnings(as.numeric(recode(prDes$devStage, "'E16'=-2; 'P2'=2; 'P6'=6; 'P10'=10", as.factor.result = FALSE)))
  #remove week 4
  miniDat<- subset(miniDat, miniDat$devStage != "4_weeks")
  return(miniDat)
  }

makexyplot <- function(x, pch = 1, cex = 1)
  {
  return(xyplot(gExp ~ age | gene, x,
  group = gType, jitter.data = TRUE,
  auto.key = TRUE, type = c('p', 'a'),
  grid = TRUE, pch = pch, cex = cex))
  }
jDat <- prepareDataAge(genes12)
makexyplot(jDat)
```
